Competing Constructions for German Measure Phrases

### 2017-03-19 Latex/article.tex
revisions of theory, more on corpus study

\section{Cognitively oriented corpus linguistics: theories and case studies}
\label{sec:cogocl}

@@

The alternation phenomenon to be discussed in this paper appears to be a straightforward case alternation at first sight.
However, without a solid knowledge of German nominal morpho-syntax, some of the more subtle distinctions to be made can be confusing.
Therefore, I introduce and illustrate the relevant constructions in some detail in this section, followed by a review of some relevant literature in Section~\ref{sec:analyses}.
I define the narrow range of syntactic constellations in which the alternation occurs and thus motivate the focus on \textit{only} this narrow range rather than the whole range of nominal constructions expressing quantities.
This entails that I also have to discuss certain constructions and factors that might be seen as relevant, but which I exclude from my own study.

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun specifying a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.%
\footnote{It would be more technically appropriate -- but also more cumbersome -- to speak of the \textit{kind noun phrase} because the embedded noun can be modified and form a full noun phrase, cf.\ immediately below and Section~\ref{sec:analyses} on the phrasal status of the constituents.}
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \citet[284]{Brems2003} considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous} (\textit{sic!}) \textit{potential for quantification} (see also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).
For illustration purposes, in the English \textit{a cup of fine coffee}, \textit{cup} is the measure noun, and \textit{coffee} is the kind noun.

Descriptively, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun comes with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{Whereas partitives are constructions denoting a proper part-of relation strictly requiring a definite kind noun as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \cite{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.}
I use \textdagger\ to denote general unacceptability, and the acceptability judgements given in (\ref{ex:intro:pseudopartitive}) are completely undisputed.
Notice that the kind noun is obligatory in the genitive and that case agreement with the measure noun is not an option (but see immediately below).
I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

When the kind noun is bare -- \ie\ when it comes without a determiner and without modifying adjectives -- the kind noun has to agree in case with the measure noun, and the genitive seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition}).
This construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.%
\footnote{Notice that the unavailability of the genitive on the kind noun can be seen as following from a quirky constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable \citep{GallmannLindauer1994,Schachtl1989}.}

The construction as in (\ref{ex:intro:narrowapposition2}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages where the PGC with the synthetic genitive is not available.%
\footnote{This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to `of' -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
While only the accusative is shown in (\ref{ex:intro:narrowapposition}), the kind noun obligatory agrees in case also with nominative and dative measure nouns.

The actual alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the NAC in (\ref{ex:intro:alternation1}) and the PGC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are also functionally and semantically equivalent.%
\footnote{Some descriptive and normative grammars take stronger positions w.\,r.\,t.\ the acceptability of these two options.
See \cite{Hentschel1993,Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As the usage and experimental data presented below should render it clear, there might be preferences under certain circumstances, but we cannot assume either construction to be unacceptable.}

\begin{exe}
  \ex\label{ex:intro:alternation}
  \begin{xlist}
    \ex[ ]{\label{ex:intro:alternation2}\gll Wir trinken [[eine Tasse]\Sub{Acc} [heißen Kaffee]\Sub{Acc}]\Sub{Acc}.\\
    we drink a cup hot coffee\\
    \glt We drink a cup of hot coffee.}
    \ex[ ]{\label{ex:intro:alternation1} Wir trinken [[eine Tasse]\Sub{Acc} [heißen Kaffees]\Sub{Gen}]\Sub{Acc}.}
  \end{xlist}
\end{exe}

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \multicolumn{1}{r}{kind NP:} & bare noun & with adjective & with determiner \\
    & [\ldots N\Subsf{meas} [N\Subsf{kind}]] & [\ldots N\Subsf{meas} [AP N\Subsf{kind}]] & [\ldots N\Subsf{meas} [D N\Subsf{kind}]] \\
    \midrule
    narrow apposition         & \CheckIt & \CheckIt &          \\
    pseudo-partitive genitive &          & \CheckIt & \CheckIt \\
  \end{tabular}
  \caption{Distribution of the NAC and PGC constructions in different NP structures}
  \label{tab:constructions}
\end{table}

\begin{enumerate}[i.]
  \item{\label{it:intro:idiot} alleged alternatives to the NAC and the PGC}
  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives}
  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} similar plural/collective construction(s)}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

First of all, (\ref{it:intro:idiot}) refers to claims found in some grammars that a generic nominative, accusative, and even dative on the kind noun are used instead of the genitive or case agreement (overviews in \citealp{Hentschel1993,Zimmer2015}).
It was shown empirically in \cite{Hentschel1993} that such variants are de facto not acceptable.
Also, in my corpus sample, they simply did not occur.
Even if they are accepted by some speakers, their extremely low frequency makes it impossible to study them using either corpus methods (Section~\ref{sec:corpusstudies}) or experimental approaches (Section~\ref{sec:externalvalidation}), and I consequently ignore them.

As for (\ref{it:intro:datsg}), a complication with neuter kind nouns in the dative is mentioned by \citet[20--22]{Zimmer2015}.
He reports a high number of occurrences of adjectives being inflected according to the \textit{weak} inflectional pattern, which is normally only used if a strongly inflected determiner precedes the adjective: \textit{kalt-en Wasser} `cold water' instead of \textit{kalt-em Wasser}.%
\footnote{\label{fn:strongweak}The \textit{strong} case and number suffixes of the adjectives are identical (with one exception) to the suffixes of determiners, and they are used when no strongly inflected determiner precedes the adjective.}
In my corpus sample (released with this article), this tendency was not nearly as clear, and there was a high number of very noisy sentences among those potentially showing this pattern.
Hence, I do not discuss these forms much, although I will offer a possible interpretation in Section~\ref{sec:analyses}.

\label{page:femininesyncretism}
With feminine kind nouns, (\ref{it:intro:femsg}) is relevant.
Whereas the singular masculine and neuter nominal subsystem is a four case system (at least for NPs with adjectives and determiners), feminine singular NPs show nominative-accusative and dative-genitive syncretisms.
Table~\ref{tab:syncretisms} shows this and the partial syncretism in the plural, where no gender distinctions are made (for an overview, see \citealp[XYZ]{Eisenberg2013a}).
This means that with dative measure nouns, the alternation is unobservable if the kind noun is feminine, and it is problematic to speak of a \textit{genitive} kind noun construction in this case.
This is taken into account in Section~\ref{sec:corpusstudies} in the model design and the discussion.

Turning to (\ref{it:intro:plurl}), readers might have noticed that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun with a collective denotation such as \textit{ein Sack kleine Äpfel\slash kleiner Äpfel} `a bag of small apples' (see Table~\ref{tab:syncretisms} for the inflection), a similar alternation between PGC and NAC can be observed.%
\footnote{\label{fn:eschenbash}On the collective interpretation of such constructions, see \cite{Eschenbach1994}.
It is, however, unfortunate that Eschenbach does not notice and discuss the morpho-syntactic distribution of PGC and NAC, even when she gets very close to listing the relevant data \citep[217]{Eschenbach1994}.}
In line with experimental results reported in \citet[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (over 92\%, cf.\ Section~\ref{sec:corpusstudies}) that the alternation cannot be analysed in the same way as in the singular case.
While this will play a role in the interpretation of the corpus findings, such plural\slash collective MNPs will not be dealt with prominently in the remainder of the paper.
A statistical model for plural kind nouns will be reported in passing in Section~\ref{sec:corpusstudies}, however.

Turning to issue (\ref{it:intro:noifl}) now, some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent', which do not have plural forms at all.%
\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
A non-inflected variant \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected variant \textit{zwei Gläser roter Wein} co-exist.
The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like any other measure noun because they enter into both the NAC as in (\ref{ex:uninflectedmeasures:a}) and the PGC as in (\ref{ex:uninflectedmeasures:b}).
See also Section~\ref{sec:analyses}.

Finally, (\ref{it:intro:other}) suggests that there are alternative ways of expressing quantification.
First of all, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

\begin{exe}
  \ex\label{ex:alternatives}
  \begin{xlist}
  \ex{\label{ex:voll}\gll   ein Glas voll von rotem Wein\\
                            a glass full of red wine\\
			    \glt a glass full of red wine}
  \ex{\label{ex:voller}\gll ein Glas {voll\slash voller} {rotem\slash roter} Wein\\
                            a glass full-of red wine\\
			    \glt a glass full of red wine}
  \ex{\label{ex:mit}\gll    ein Glas mit rotem Wein\\
                            a glass with red wine\\
			    \glt a glass with red wine in it}
  \end{xlist}
\end{exe}

These construction have very specific properties (the construction with \textit{voller} are discussed in \citealp{Zeldes2018}, the construction with \textit{mit} is discussed in \citealp{Bhatt1990}), they are not semantically equivalent to NAC and PGC, and -- most importantly -- they are only available for a small subset of measure nouns.
Subtle interactions between these constructions and NAC\slash PGC cannot be excluded, especially because the former might act as a (partial) alternative and thus reduce the count of exemplars of the NAC, the PGC, or both for container measure nouns.
There are no clear and motivated hypotheses about such interactions, however.
Nevertheless, the corpus-based models in Section~\ref{sec:corpusstudies} are specified in a way that they could detect preferences for either NAC or PGC for those nouns for which the alternative constructions illustrated in (\ref{ex:alternatives}) is available.

This concludes the descriptive overview of the alternation.
I have shown that there is an alternation between two very specific measure noun constructions under quite specific syntactic conditions, and that the two constructions can be distinguished clearly by looking at the case of the kind noun.
It should be kept in mind that this paper is about this specific alternation and not about a possibly broader range of nominal constructions expressing quantification or measure.
I turn to some more theory-oriented discussion in the next section.

@@

This section reviews some existing analyses of the PGC-NAC alternation or related issues, and I develop my own analysis and the appropriate hypotheses for the studies in Sections~\ref{sec:corpusstudies} and \ref{sec:externalvalidation}.
While the constructions and the alternation are often covered in descriptive grammars such as \citet[XYZ]{Eisenberg2013a}, the available linguistic literature contains little specific discussion of the situation in German.
In The remainder of this section, I discuss as factors that might influence the alternation:
degrees of grammaticalisation of the measure noun, the different syntactic structures of the variants and their syntactic prototypes, and potential effects of semantic prototypicality.

Turning to grammaticalisation, it is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).%
\footnote{Notice that the description of German in \citet[547--549]{Koptjevskaja2001} is vague and partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
Her example (29b) from \citet[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun (hence, PGC).

It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \pageref{page:femininesyncretism} and an ambiguous case requirement of the preposition \textit{troz}, hence an instance of NAC.
Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
}
The alternating constructions in German are both clearly pseudo-partitives, but the grammaticalisation paths assumed by \cite[esp.\ 526--530]{Koptjevskaja2001}, which I interpret freely here, are clearly pertinent.
The development usually begins with constructions involving two referential nouns (not necessarily forming a single NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the case} represents the first stage of a development wherein the measure noun already tends to loose some semantic content.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers as in English \textit{a lot}.
Since in German, the PGC is clearly the older variant \citep{Zimmer2015} with the overt \textit{quantity-of} marker in the form of the genitive suffix and at least the potential to form a true partitive (if the kind noun is definite), which the NAC lacks, we can expect more strongly grammaticalised measure nouns to prefer the NAC.
Since clearly non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should favor the NAC according to this hypothesis, the effect can be detected by including information about measure noun classes in the corpus-based models in Section~\ref{sec:corpusstudies}.

On the other hand, there are case studies like \cite{Brems2003,DeclerckBrems2016} where syntactic evidence for grammaticalisation plays a prominent role.
In English -- \cite{Brems2003} argues based on a view from \cite{Langacker1991} -- strongly lexicalised and thus semantically bleached measure nouns turn from nouns into quantifiers, leading to a syntactic reanalysis of the NP, wherein the kind noun becomes the head.
The argument is supported by convincing syntactic evidence, for example subject-verb agreement.
In German MNPs, the measure noun always remains the syntactic head, however (see immediately below for references), and there is no similar clear criterion to detect grammaticalisation.
The following analysis of the morpho-syntax of German MNPs will however, provide hypothetical cues along similar lines.

I therefore turn to the MNP-internal syntax, which will show itself to be highly revealing.
It cannot be understood properly without taking into account the strong\slash weak inflection patterns of adjectives described in Section~\ref{sec:descriptive} (especially Footnote~\ref{fn:strongweak}).
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretistic \textit{weak} pattern.%
\footnote{The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners, which is called the \textit{strong} pattern (see Table~\ref{tab:syncretisms}).
Thus, the adjectives in the target constructions have properties of adjectives as well as determiners in a sense:
they are lexical adjectives and function as attributive modifiers, but they are inflected as determiners and are the leftmost element in the NP.
This unusual double nature of adjectives in NPs without a determiner leads to a plausible interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase as a determiner, they have to use the PGC because, if there is a determiner, the PGC is the only option.
When they classify the adjective as an adjective, the kind NP has no determiner, and the NAC is the only option.%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the GPC as a DP and in the NAC as an NP.
Within the confines of her framework, this is the appropriate was of encoding the relevant generalisation.}
In a framework that assumes similarity-based probabilistic classification, such as Prototype Theory, we can now regard the PGC with a determiner in the kind noun phrase as instantiating the PGC prototype and the NAC with just a bare kind noun as the NAC prototype.
The choice of the alternating variants is then expected to be influenced by the degree of similarity to either of these prototypes.

The situation is made difficult by the fact that the semantic differences between the variants are at best subtle, which means the prototypes are not associated with nice and clearly detectable differences in meaning.
However, both the kind noun lemmas and the measure noun lemmas probably appear with different relative frequencies in the NAC with a bare kind noun and the PGC with a determiner inside the kind noun phrase (in part possibly due to their meaning and usage).
Their relative frequencies in these non-alternating configurations should be indicative of their preference to appear in one of the two variants.
In other words, lemmas themselves might have preferences for one prototype or the other.

%Related to the , dative measure nouns favor the PGC more than nominative and accusative (also \citealp{Hentschel1993,Zimmer2015}).
%As an example, \textit{mit einem Stück frischen Brots} `with a piece of fresh bread' (PGC) would be preferred over \textit{mit einem Stück frischem Brot} (NAC).
%I propose

Furthermore, the grammaticalisation path as described above leads from NPs denoting highly individuated objects standing in a \textit{separated-from} relation to a construction with a \textit{part-of} meaning (still available in the prototypical PGC) to a construction with a more diffuse \textit{quantity-of} relation (characteristic of the prototypical NAC).
The best evidence for the relevance of this factor in an ongoing grammaticalisation process is that plural kind nouns almost always trigger PGC and virtually never NAC (Section~\ref{sec:descriptive}).
More indirectly, plurality of the measure noun could favor the PGC over the NAC, and exact quantification in the form of cardinals even more so.%
\footnote{The effect of the plural was empirically shown but not explained in \cite{Hentschel1993}.}
As a highly illustrative example (using a measure noun which has virtually developed two variants, one of which grammaticalsed to a much higher degree), \textit{ein Haufen französische Watte} `a heap\slash lot of French cotton' (NAC) has essentially two uses.
It can be used to denote an unspecific amount of cotton in the sense of \textit{a lot of French cotton} in situations when the exact quantity of cotton is not of interest.
Under the appropriate circumstances, it could also be used to denote a highly specific quantity of cotton, for example when a cotton merchant always sells her French cotton in \textit{heaps} weighing two stone each.
The former is highly restricted for \textit{zwei Haufen französischer Watte} `two heaps/\textdagger lots of French cotton' (GPC).
Such an NP is less likely to be used in situations where the exact quantity of cotton is irrelevant, even if there are actually two heaps of cotton.
In other words, a precisely quantified measure noun leads to a higher individuation (the creation of portions) of the substance denoted by the kind noun, which is more appropriate for the PGC prototype.%
\footnote{In the most extreme, some measure nouns cannot be pluralised, such as \textit{eine Menge} `a lot of'.
Also note that the presentation of the examples in NAC and PGC is suggestive.
Native speakers will confirm that both NPs sound highly acceptable in both GPC and NAC.
It was made clear from the beginning and throughout that the preferences are indeed subtle.}

Finally, it is claimed that the PGC is more typical of higher registers or even exclusive to written language (see \citealp[320--323]{Hentschel1993}, where she also refers to descriptive grammars).
This is not surprising in as much as the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial and vernacular variants of German, and thus overrepresented to the same degree as colloquial, vernacular, and spoken variants are overrepresented in most corpora of written German.
There is no reason to assume that such preferences for cannot be part of the construction prototypes.
In the corpus study, the alternation tendencies controlled by this complex factor related to register, medium, and lectal variation will be tested using certain very simple proxies.

To summarise the discussion and core hypotheses, it is expected that:

\begin{enumerate}[i.]
  \item classes of more clearly grammaticalised measure nouns favor the NAC,
  \item the relative frequencies with which measure noun lemmas and kind noun lemmas appear in the prototypical (unequivocal) PGC and NAC are positively correlated with the relative frequency the PGC and the NAC are chosen in the alternating case,
  \item MNPs with masculine dative kind nouns disfavour the GPC comparatively less
  \item pluralic measure nouns and measure nouns quantified by cardinals favor the PGC variant
  \item .
\end{enumerate}

In the next section, I report the corpus study that was designed to test these hypotheses on usage data.

Before I turn to the concrete sampling procedure, a few words are in order regarding the choice of the corpus.
I used the German \textit{Corpus from the Web} (COW) DECOW14(A) (\citealp{SchaeferBildhauer2012full,Schaefer2015b} and \citealp{BiemannEa2013,SchaeferBildhauer2013} for overviews of web corpora in general and the methodology of their construction), which contains 21 billion tokens, for two main reasons.%
\footnote{The corpora are made available for free at \url{https://www.webcorpora.org}.
At the time of this writing, the DECOW16 corpus had already been released.}
First, the external validity of the study is increased through a higher heterogeneity of the sample \citep[30]{MaxwellDelaney2004}, and the DECOW14 corpus has clearly a much more heterogeneous composition compared to the only other very large corpus of German, the DeReKo \citep{KupietzEa2010} of the Institute for the German Language (IDS), which contains almost exclusively newspaper texts.%
\footnote{It was shown in \cite{W16-2601} that, for example, the spread of topics is much smaller in DeReKo compared to DECOW14.}
Second, it was already mentioned that normative grammars disagree on whether both NAC and PGC are grammatical in the case at hand.
Thus, newspaper text or any other text that conforms strongly to normative standards might not represent the alternation phenomenon fully (and without bias) because authors and proof-readers might favour one alternative or the other.
Web corpora, on the other hand, contain at least some amount of non-standard language from forums and similar sources.
For these or similar reasons, COW corpora have been used in a number of peer-reviewed publications, for example \cite{VanGoethemHiligsmann2014,VanGoethemHuening2015,MuellerS2014,Schaefer2016c,SchaeferSayatz2014,SchaeferSayatz2016,StefanowitschFlach2016,Zimmer2015}. 
Therefore, DECOW14 can be considered the obvious choice for this study.

In Section~\ref{sec:germanmeasurenps} it was mentioned that there are vague intuitions about lemma-specific preference effects w.\,r.\,t.\ the kind noun.
Also, degrees of grammaticalisation of the measure noun might influence the choice of one of the constructions.
Therefore, it was necessary to obtain sample in which all (or at least most of the highly frequent) actually occurring combinations of kind noun and measure noun were represented.
This section describes the three-stage bootstrap process that was used to obtain such a sample, which was then used to estimate the influences of various contextual factors on the alternation (cf.\ Section~\ref{sec:corpusstudies}).
The bootstrap process takes into account the imperfection of automatically generated annotations in corpora.

Automatic linguistic annotation (for lemma, part-of-speech, morphological information, etc.) is never perfect, especially if the material deviates from the standard as understood by the creators of the tools used for the annotation (cf. \citealp{BeisswengerEa2016,GiesbrechtEvert2009} and chapter 4 of \citealp{SchaeferBildhauer2013}).
For examining grammatical alternations, relying on existing annotations is therefore unwise because there is no systematic way of assessing the accuracy of the annotation tools on the target constructions.
Therefore, I did not rely on the annotations for part-of-speech, noun phrase chunks, or noun case available in DECOW14 wherever it was crucial to sample with a near-perfect recall.%
\footnote{The recall is the percentage of retrieved data relative to the data that should have been found.
If, for example, the corpus contains one hundred targets and a query retrieves only 80 of them, the recall is 80\% (or 0.8).}

Instead, I bootstrapped a sample which most likely contained all reasonably frequent combinations of measure nouns and mass nouns, and which was almost perfectly clean (\ie\ it did contain virtually no erroneous non-target results).
The bootstrapping consisted of three steps:\\

\begin{enumerate}[i.]
  \item\label{enum:data:step1} bootstrapping a list of the one hundred most frequent mass nouns,
  \item\label{enum:data:step2} bootstrapping a list of all measure nouns with which the mass nouns co-occur in the unambiguous measure construction with a determiner, and
  \item\label{enum:data:step3} sampling the target constructions by querying each combination of mass noun and measure noun found in step \ref{enum:data:step2}.
\end{enumerate}

In step \ref{enum:data:step1}, I sampled a list of all nouns in the DECOW14A01 corpus sorted by their token frequency and manually went through it from the most frequent noun downwards, selecting the first one hundred mass nouns that occurred in the list.%
\footnote{DECOW14A01 is the first slice (roughly a twentieth) of the complete DECOW14A corpus.
It contains just over one billion tokens.
The creators assigned the documents randomly to the slices, such that each slice can be considered representative of the whole corpus \cite{Schaefer2015b}.}
For this step, I relied on the noun part-of-speech label because for frequent nouns, high tagging accuracy can be expected, especially given the fact that German nouns are always spelled with an initial capital letter.
Abstract nouns which partially behave like mass nouns (like \textit{Spaß} `fun’ or \textit{Gefahr} `danger’) were excluded because they are usually not quantified in the same way as concrete mass nouns.%
\footnote{They behave like mass nouns mostly in that they occur without a determiner in the singular.}
The hundredth selected mass noun was \textit{Schmuck} `jewelry’, which is the 3,054th most frequent noun in the original frequency list.

This resulting list of mass nouns was used in step \ref{enum:data:step2} to bootstrap a list of measure nouns co-occurring with the mass nouns from DECOW14A01.
In order to generate this list, I utilized the fact that a direct sequence of two bare nouns very often instantiates the bare-noun NAC if the second noun is a mass noun.
In other words, I searched for all combinations [N\Sub{1} N\Sub{2}] with N\Sub{2} instantiated as a mass noun lemma.
Then, the resulting lists of noun--noun combinations were each sorted by frequency in descending order and sieved manually to remove erroneous hits.%
\footnote{In order to speed up the procedure without losing much material, I stopped when I had reached noun–noun combinations with an absolute frequency of 1 if twenty combinations for the respective mass noun (with a frequency of one or higher) had already been found.}
The result was a list of 2,365 individual combinations of measure nouns and mass nouns.

In step \ref{enum:data:step3}, each of these 2,365 noun–noun combinations was queried in the target construction individually in each of the first ten slices of DECOW14 (roughly 10 billion tokens).
The scripts generating the actual queries were crafted so as to make sure that adjectives and nouns were searched in the appropriate forms.
(\ref{ex:data:query01}) shows an example query for a neuter mass noun like \textit{Wasser} (`water’) which finds relevant sequences such as the ones shown in (\ref{ex:data:query02}), which all mean `litre of clear water’.
The versions with the genitive singular \textit{Liters} and the dative plural \textit{Litern} are not shown in (2) for space reasons but are also found by the query in (1).
In terms of context, single sentences containing the target were exported.

To reduce the sample size for the manual annotation process, the final concordance was sampled from the results of these 2,365 queries.
Since the mass nouns in the sample were distributed according to the usual power law, I used all hits for mass nouns occurring less than one hundred times, but randomly sampled one hundred hits for each mass noun that occurred one hundred or more times.
The final sample contained 7,547 sentences.
Given the careful bootstrapping and sampling procedure described in this section, we can be highly sure that it contains all relevant and reasonably frequent noun–noun combinations in the target construction.

\subsection{A prelude to the statistical analysis: There are no right statistics!}
\label{sec:rightstatistics}

@@

>>>>

\section{Cognitively oriented corpus linguistics: theories, case studies, and methods}
\label{sec:cogocl}

\subsection{Corpora and cognition}

@@

@@

\subsection{A word on statistical analysis}
\label{sec:rightstatistics}

@@

The alternation phenomenon to be discussed in this paper appears to be a straightforward case alternation at first sight.
However, without a solid knowledge of German nominal morpho-syntax, some of the more subtle distinctions that have to be made in its analysis can be confusing.
Therefore, I introduce and illustrate the relevant constructions in some detail in this section.
The narrow range of syntactic constellations in which the alternation occurs is described, and the focus on \textit{only} this narrow (range rather than, for example, the whole range of nominal constructions expressing quantities) is motivated.
This entails that I also have to discuss certain constructions and factors that might be seen as relevant, but which I exclude from my own study.

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun specifying a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \citet[284]{Brems2003} considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous} (\textit{sic!}) \textit{potential for quantification} (see also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).
For illustration purposes, in the English \textit{a cup of fine coffee}, \textit{cup} is the measure noun, and \textit{coffee} is the kind noun.

In the case at hand, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun forms an NP with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{Whereas partitives are constructions denoting a proper part-of relation strictly requiring a definite kind noun as in \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities and contain indefinite kind nouns as in \textit{a sip of wine}.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \cite{Hentschel1993} while some realise the difference and at least mention it \citep{Eschenbach1994,GallmannLindauer1994,Loebel1989,Zimmer2015}.}
I use \textdagger\ to denote general unacceptability, and the acceptability judgements given in (\ref{ex:intro:pseudopartitive}) are completely undisputed.
Notice that the kind noun is obligatorily in the genitive and that case agreement with the measure noun is not an option (but see immediately below).
I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

When the kind noun is bare -- \ie\ when it comes without a determiner and without modifying adjectives -- the kind noun has to agree in case with the measure noun, and the genitive seen in the PGC is not acceptable, see (\ref{ex:intro:narrowapposition}).
This construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.%
\footnote{Notice that the unavailability of the genitive on the kind noun can be seen as following from a rather quirky constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable \citep{GallmannLindauer1994,Schachtl1989}.}

The construction as in (\ref{ex:intro:narrowapposition2}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages in which the PGC with the synthetic genitive is not available.%
\footnote{This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to `of' -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
While only an accusative is shown in (\ref{ex:intro:narrowapposition}), the kind noun obligatory agrees in case also with nominative and dative measure nouns.

The actual alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the NAC in (\ref{ex:intro:alternation1}) and the PGC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are by-and-large functionally and semantically equivalent.
However, Section~\ref{sec:analyses} is devoted to developing hypotheses about subtle differences between them.%
\footnote{Some descriptive and normative grammars take stronger positions w.\,r.\,t.\ the acceptability of the two options.
See \cite{Hentschel1993,Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As the usage and experimental data presented below should render it clear, there might be preferences under certain circumstances, but we cannot assume either construction to be unacceptable.}

\begin{exe}
  \ex\label{ex:intro:alternation}
  \begin{xlist}
    \ex[ ]{\label{ex:intro:alternation1}\gll Wir trinken [[eine Tasse]\Sub{Acc} [heißen Kaffee]\Sub{Acc}]\Sub{Acc}.\\
    we drink a cup hot coffee\\
    \glt We drink a cup of hot coffee.}
    \ex[ ]{\label{ex:intro:alternation2} Wir trinken [[eine Tasse]\Sub{Acc} [heißen Kaffees]\Sub{Gen}]\Sub{Acc}.}
  \end{xlist}
\end{exe}

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \multicolumn{1}{r}{kind NP is:} & bare noun NP & NP with adjective & NP with determiner \\
    & [\ldots N\Subsf{meas} [N\Subsf{kind}]] & [\ldots N\Subsf{meas} [AP N\Subsf{kind}]] & [\ldots N\Subsf{meas} [D N\Subsf{kind}]] \\
    \midrule
    narrow apposition         & \CheckIt & \CheckIt &          \\
    pseudo-partitive genitive &          & \CheckIt & \CheckIt \\
  \end{tabular}
  \caption{Distribution of the NAC and PGC constructions in different NP structures}
  \label{tab:constructions}
\end{table}
